---
title: "Imputation Pipeline: QUALITY CONTROL"
output: html_document
author: "NM - ODAP - ICO"
date: "`r Sys.Date()`"
params:
  dataDir:
    label: "dataDir"
    value: ""
  plinkFile:
    label: "plinkFile:"
    value: ""
---


# 0 data preparation

```{r setup, include=FALSE}

library(knitr)

setwd("./1_QC")

dataDir <- "./QC/"

# Replace with the names of the files of interest:
plinkFile <- "plink_raw_data_file_name"

outputDir<-"./QC/"

# read samples to mark them
fam <- read.table(paste0(dataDir, plinkFile, ".fam") ,stringsAsFactors = FALSE, header = FALSE, sep = "")
rownames(fam) <- fam[,2]

```

<br>

Filter out SNPs that don't pass quality control.

Mark or remove samples that don't pass quality control.

<br>


## b Requisites:

* R 3.5.0
* plink 1.9
* AIMS_SNPs.txt file must be in the code folder
* Data must be in plink format (bed + bim + fam)

<br>  


```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

cat("\n")
cat(paste0("Quality control for ",outputDir,plinkFile," data."))
cat("\n")

cat("\n")
cat(paste0("There are ",nrow(fam)," samples in ",plinkFile," data."))
cat("\n")

bim <- read.table(paste0(outputDir, plinkFile,".bim") ,stringsAsFactors = FALSE, header = FALSE, sep = "")


cat("\n")
cat(paste0("There are ",nrow(bim)," SNPs in ",plinkFile," data."))
cat("\n")

```


# a) Missing SNPs:

Filter SNPs with 5% of missings

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# missing proportion
system(paste0("plink --bfile ", dataDir, plinkFile, " --missing --out ", outputDir, "missings"))
missings <- read.table(paste0(outputDir, "missings.lmiss"), header=TRUE)


## SNPs to exclude 
exclude_miss <- subset(missings, F_MISS > 0.05)
write.table(exclude_miss[,2], file = paste0(outputDir, "snp_exclude_miss.txt"), col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")

# filter SNPs with missings
system(paste0("plink --bfile ", dataDir, plinkFile, " --exclude ", outputDir, "snp_exclude_miss.txt --make-bed --out ", outputDir, plinkFile, "_miss"))

cat("\n")
cat(paste0("There are ",nrow(exclude_miss)," SNPs with more than 5% of missings. The complete list of SNPs to remove will be saved in SNPsmissings.txt"))
cat("\n")


## print plot
histplot <- hist(missings$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num SNPs", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "removed SNPs", adj = 0, col = "darkred")
box(lwd = 2)

## save plot
png(paste0(outputDir, "SNPmissings.png"), res = 200, 1500, 1500)
histplot <- hist(missings$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num SNPs", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "removed SNPs", adj = 0, col = "darkred")
box(lwd = 2)
dev.off()

# save excluded SNPs
write.table(exclude_miss, file = paste0(outputDir, "SNPsmissings.txt"), row.names = FALSE, quote = FALSE, sep = "\t")

```

<br>

# b) Heterozigosity and Missing samples: 


```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# Heterozygosity rate
system(paste0("plink --bfile ", outputDir, plinkFile, "_miss --het --out ", outputDir, "heterozygotsity"))
hetero <- read.table(paste0(outputDir, "heterozygotsity.het"), header = TRUE)
hetero$het <- (hetero[,5] - hetero[,3]) / hetero[,5]

# Cheking mean+-4sd
meanup <- mean(hetero$het) + 4 * sd(hetero$het)
meandown <- mean(hetero$het) - 4 * sd(hetero$het)



# Missings - CALL RATE >90 (F_MISS >0.1)
system(paste0("plink --bfile ", outputDir, plinkFile, "_miss --missing --out ", outputDir, "missings"))
missInd <- read.table(paste0(outputDir, "missings.imiss"), header=TRUE)

# plot

plot(log(hetero$het),log(missInd$F_MISS),pch=20,main=paste0("HZ>",round(meanup,2)," (n=",sum(hetero$het>meanup),")\nHZ<",round(meandown,2)," (n=",sum(hetero$het<meandown),")"))
abline(v=log(meanup),col="darkred")
abline(v=log(meandown),col="darkred")
abline(h=log(0.1),col="darkred")
box(lwd = 2)

png(paste0(outputDir, "HZvsMISS_4sd.png"), res = 200, width=1500, height=1500)
plot(log(hetero$het),log(missInd$F_MISS),pch=20,main=paste0("HZ>",round(meanup,2)," (n=",sum(hetero$het>meanup),")\nHZ<",round(meandown,2)," (n=",sum(hetero$het<meandown),")"))
abline(v=log(meanup),col="darkred")
abline(v=log(meandown),col="darkred")
abline(h=log(0.1),col="darkred")
box(lwd = 2)
dev.off()



####################################################################################################
```

## b.1.- Missing samples

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}
# remove missing samples <90 CALL RATE

cat("\n")
cat(paste0("**b.1.- Missings:**"))
cat("\n")

## Samples to mark
write.table(missInd, paste0(outputDir,"sample_callrate_table.txt" ))
exclude_miss <- subset(missInd, F_MISS > 0.1)


torm<-fam[fam$V2 %in% exclude_miss$IID,]

cat("\n")
cat(paste0("There are ",nrow(torm)," samples with more than 10% of missings."))
cat("\n")


## print plot
histplot <- hist(missInd$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num samples", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.11, max(histplot$counts)/2, "removed samples", adj = 0, col = "darkred")
box(lwd = 2)

## save plot
png(paste0(outputDir, "SAMPLEmissings.png"), res = 200, 1500, 1500)
histplot <- hist(missInd$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num samples", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "removed samples", adj = 0, col = "darkred")
box(lwd = 2)
dev.off()





###############

# mark samples
fam$miss_samp <-"No"
fam[torm$V2,"miss_samp"]<-"Low_CR_S"


write.table(torm[,1:2], file = paste0(outputDir,"/rm_Missing_CR_90_samples.txt"), col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")

system(paste0("plink --bfile ",outputDir, plinkFile,"_miss --remove ",outputDir,"/rm_Missing_CR_90_samples.txt --make-bed --out ",outputDir,plinkFile,"_miss_sample"))

```

## b.2.- Heterozigosity:

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

####################################################################################################

cat("\n")
cat(paste0("**b.2.- Heterozigosity:**"))
cat("\n")

hzdown<-hetero[hetero$het<meandown,c(1,2,7)]

cat("\n")
cat(paste0("The lower heterozygosity threshold (mean-4sd) is ",round(meandown,2),". There are ",nrow(hzdown)," samples that are outside this threshold."))
cat("\n")

kable(hzdown,row.names=FALSE)

hzup<-hetero[hetero$het>meanup,c(1,2,7)]

cat("\n")
cat(paste0("The upper heterozygosity threshold (mean+4sd) is ",round(meanup,2),". There are ",nrow(hzup)," samples that are outside this threshold."))
cat("\n")

kable(hzup,row.names=FALSE)

cat("\n")
cat(paste0("Only samples with high HZ will be removed."))
cat("\n")


# mark samples
fam$ExcludeHeteroz <-"No"
fam[hzdown$IID,"ExcludeHeteroz"]<-"LowHZ"
fam[hzup$IID,"ExcludeHeteroz"]<-"HighHZ"

table(fam$ExcludeHeteroz, fam$miss_samp)
  #           Low_CR_S No
  # HighHZ       17    0
  # No           64    1146

# tots els HighHZ presenten call rate <90. ja hena estat eliminats

torm<-fam[fam$ExcludeHeteroz=="HighHZ",]
write.table(torm[,1:2], file = paste0(outputDir,"/rm_HighHZ_samples.txt"), col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")

system(paste0("plink --bfile ",outputDir, plinkFile,"_miss_sample --remove ",outputDir,"/rm_HighHZ_samples.txt --make-bed --out ",outputDir,plinkFile,"_HZ"))


unlink(paste0(outputDir,"*.log"))
unlink(paste0(outputDir,"*.hh"))
unlink(paste0(outputDir,"*miss*"))

```

<br>

# c) Sex concordance:

Remove samples with no sex concordance

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# sex check
system(paste0("plink --bfile ",outputDir,plinkFile,"_HZ --check-sex --out ",outputDir,"sexConc"))

unlink(paste0(outputDir,"*.log"))
unlink(paste0(outputDir,"*.hh"))

sex<-read.table(paste0(outputDir,"sexConc.sexcheck"),sep="",stringsAsFactors = FALSE,header=TRUE)

# filter in samples with no concordance
sex <- sex[sex$STATUS!="OK" & sex$SNPSEX!=0 & sex$PEDSEX!=0,]

# mark samples
fam$ExcludeSex<-FALSE
fam[as.character(sex[,2]),"ExcludeSex"]<-TRUE


torm<-fam[fam$ExcludeSex,]
write.table(torm[,1:2], file = paste0(outputDir,"/rm_Sex_samples.txt"), col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")

system(paste0("plink --bfile ",outputDir, plinkFile,"_HZ --remove ",outputDir,"/rm_Sex_samples.txt --make-bed --out ",outputDir,plinkFile,"_QC"))


cat("\n")
cat(paste0("There are ",sum(fam$ExcludeSex)," samples with no sex concordance."))
cat("\n")

if(sum(fam$ExcludeSex)>0) print(kable(fam[fam$ExcludeSex,c(1,8)],row.names=FALSE))

```

<br>

# d) Duplicates and relatedness (PLINK pi_hat > 0.8):

Mark duplicated and related samples

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# check relatedness
system(paste0("plink --bfile ", outputDir, plinkFile, "_QC --genome --out  ", outputDir, "duplicates"))
dups <- read.table(paste0(outputDir, "duplicates.genome"), header = TRUE)

# filter in related or duplicated individuals
dups <- dups[dups$PI_HAT > 0.8,]

# mark samples
fam$ExcludeDups <- "No"
if(nrow(dups)>0){
  for(dp in 1:nrow(dups)){
    fam[as.character(dups[dp,2]),"ExcludeDups"] <- as.character(dups[dp,4])
    fam[as.character(dups[dp,4]),"ExcludeDups"] <- as.character(dups[dp,2])
  }
}

cat("\n")
cat(paste0("There are ",sum(fam$ExcludeDups!="No")," duplicated or related samples."))
cat("\n")

#print(kable(dups[,c(2,4,10)],row.names=FALSE))

# remove some files
unlink(paste0(outputDir,"*.log"))
unlink(paste0(outputDir,"*.hh"))


```

<br>



# e) Hardy-Weinberg equilibrium:

```{r, eval=FALSE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# NOT EXECUTED!!! EVAL=FALSE!!!!

### e) Hardy-Weinberg equilibrium:

#Filter SNPs that have a HWE p-value < 1e-04

# check HWE
system(paste0("plink --bfile ", outputDir, plinkFile, "_QC --hardy --out ", outputDir, "hwe"))
hwe <- read.table(paste0(outputDir, "hwe.hwe"), header = TRUE)
hwe <- hwe[hwe$TEST=="UNAFF",] # only if there are cases/controls in the data

## SNPs to exclude
exclude_hwe <- subset(hwe, P<1e-04)

cat("\n")
cat(paste0("There are ",nrow(exclude_hwe)," snps that don't pass the HWE.  The complete list of SNPs to remove will be saved in SNPsHWE.txt"))
cat("\n")

# Remove SNPs
write.table(exclude_hwe[,2],file = paste0(outputDir, "snp_exclude_hwe.txt"), quote = FALSE, col.names = FALSE, row.names = FALSE)
system(paste0("plink --bfile ", outputDir, plinkFile, "_QC --exclude ", outputDir, "snp_exclude_hwe.txt --make-bed --out ", outputDir, plinkFile, "_QC"))


# remove some files
unlink(paste0(outputDir,"snp_exclude_hwe.txt"))
unlink(paste0(outputDir,"*.hh"))
unlink(paste0(outputDir,"*.log"))
unlink(paste0(outputDir,"hwe.hwe"))
unlink(paste0(outputDir, "*HZ*"))

# save removed SNPs 
write.table(exclude_hwe, file = paste0(outputDir, "rm_HWE_SNPs.txt"), row.names = FALSE, quote = FALSE, sep = "\t")



```

<br>

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

bim <- read.table(paste0(outputDir, plinkFile, "_QC.bim") ,stringsAsFactors = FALSE, header = FALSE, sep = "")

cat("\n")
cat(paste0("Finally, ",nrow(bim)," SNPs remained for the analysis."))
cat("\n")

write.table(fam, file = paste0(outputDir, "marked_samples.txt"), sep = "\t", quote = FALSE, row.names = FALSE)

```

<br>


# g) Genetic population statification:

Mark samples outside the ancestry pupulation cluster of data using AIMS (Ancestry-informative marker SNPs) (abs(PC1) > 0.1 & abs(PC2) > 0.1). 

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

system(paste0("plink --bfile ", outputDir, plinkFile, "_QC --extract ", dataDir, "AIMS_SNPs.txt --make-bed --out  ", outputDir, plinkFile, "_AIMS"))
system(paste0("plink --bfile ", outputDir, plinkFile, "_AIMS --pca --out  ", outputDir, "pca"))

pca <- read.table(paste0(outputDir, "pca.eigenvec"), header = FALSE)

## print plot
plot(pca[,3], pca[,4], pch = 20, xlab = "PCA 1", ylab = "PCA 2", main = plinkFile)

# save plot
png(paste0(outputDir, "Ancestry.png"), res = 200, 1500, 1500)
plot(pca[,3], pca[,4], pch = 20, xlab = "PCA 1", ylab = "PCA 2", main = plinkFile)
dev.off()


##########################################

# PCA in our study
pca<- read.table(paste0(outputDir, "pca.eigenvec"), header = FALSE)

pca$V1<-NULL
colnames(pca)<-c("ID",paste0("PC",1:5))


# european variable
pca$European<-"European"
pca$European[pca$PC2>(-0.005)]<-"Latino"
pca$European[pca$PC2>0.03]<-"Asian"
pca$European[pca$PC1<(-0.01)]<-"African"


table(pca$European)
# African European   Latino 
#       24       17      171 

covar<-pca




png(paste0(outputDir, "Ancestry_colour.png"), res = 200, 1500, 1500)
plot(covar$PC1, covar$PC2, main="",pch=20,cex=1.2,col="black",xlab="PCA 1", ylab="PCA 2",las=1,cex.lab=1.2,cex.axis=1.2)
points(covar$PC1[covar$European=="Asian"], covar$PC2[covar$European=="Asian"],cex=1.2,pch=20,col="red")
points(covar$PC1[covar$European=="African"], covar$PC2[covar$European=="African"],cex=1.2,pch=20,col="dodgerblue")
points(covar$PC1[covar$European=="Latino"], covar$PC2[covar$European=="Latino"],cex=1.2,pch=20,col="green")
legend("topleft",c("European","Asian","African","Latino"),col=c("black","red","dodgerblue","green"),pch=20,bty="n",cex=2)
box(lwd=2)
dev.off()


#############################################################


## Mark samples
fam$ExcludePCA <- "No"

write.table(fam, file = paste0(outputDir, "marked_samples.txt"), sep = "\t", quote = FALSE, row.names = FALSE)

unlink(paste0(outputDir, plinkFile, "_AIMS.*"))
unlink(paste0(outputDir, "pca.log"))
unlink(paste0(outputDir, "pca.eigenval"))

```



#### Output files:

The output files of the QC anlysis are:

* Three 'good quality' SNPs and samples filtered in plink files (plinkFile_QC.bed; plinkFile_QC.bim and plinkFile_QC.fam)

* Two plots of missigness (SNPmissings.png and SAMPLEmissings.png)

* The list of SNPs with more than 5% of missings (SNPmissings.txt)

* Heterozigosity file (heterozygotsity.het)

* Heterozigosity vs Call rate plot (HZvsMISS_4sd.png)

* Removed samples (rm_HighHZ_samples.txt, rm_Missing_samples.txt, rm_Sex_samples.txt)

* Sex concordance file (sexConc.sexcheck)

* Duplicates file (duplicates.genome)

* A txt with the samples marked (marked_samples.txt). fam format with some extra columns.

<br>  
<br>  


